Frame Competition - Assessing the Impact of News Framing on Parties' Electoral Performance with Document Embeddings

### 2022-12-08 paper/Manuscript.Rmd
Add plots
Preliminary Analyses, update Abstract

---
title: "Frame Competition"
author: "Nicolai Berk"
date: "`r Sys.Date()`"
output:
  bookdown::pdf_document2:
    toc: false
    fig_caption: yes
    includes:
      in_header: preamble.tex
subtitle: Manuscript prepared for the OX|BER workshop, Berlin, July 14th-15th.
fontsize: 12pt
geometry: margin=1.5in
abstract: "Issue ownership theory assumes that the salience of specific issues is always more or less beneficial to a given party. I introduce the concept of 'frame competition' and argue that a given issue can be more or less beneficial for parties dependent on the dominant framing. I test this argument using data from the German 2017 Bundestag campaign, assessing the similarity of media coverage on migration to parties' migration framing with topic vectors. Using VAR models and an instrumental variable approach, I show whether higher frame similarity is associated with better performance in the polls. The findings broaden our understanding party competition and underline possibilities for political change, as well as parties' capability to compete, even when the issue agenda is given."
bibliography: C:/Users/nicol/Dropbox/Studium/BibTex/PhD.bib
---

@@

knitr::opts_chunk$set(echo = F, message = F, warning = F, cache = F, fig.height=6, fig.width=10, fig.align='center', eval = T)

@@

@@

One of the most prominent explanations for parties' electoral success in scholarly and public debate is the issue agenda at a given election: the more prominent an issue was among political parties, news media, and voters, the more it would influence the latter's electoral decision making. Certain issues are therefor more favorable to given parties and parties will try to focus on 'their' issues in public communication [@Budge1983; @Petrocik1996; @Walgrave2012]. In extension, it is assumed that the main form of electoral competition in contemporary Western European democracies is *issue competition*, in the sense that parties compete to bring 'their' issue on the agenda [@GreenPedersen2019].

@@

@@

@@

afd_poll_plot <- 
  de_polls_df %>% 
  mutate(date = as.Date(date)) %>% 
  filter(date < as.Date("2016-07-01"),
         date >= as.Date("2015-01-01")) %>% 
  mutate(week = lubridate::floor_date(date, "week")
         # , post = date >= as.Date("2016-01-01")
         ) %>% 
  group_by(week
           # , post
           ) %>% 
  summarise(parties.AfD = mean(parties.AfD)) %>% 
  ggplot(aes(week, parties.AfD
             # , group = post
             )) +
  geom_point() +
  geom_smooth(method = "gam") +
  # geom_vline(xintercept = as.Date("2015-09-01"), col = "red", lty = 2) +
  theme_minimal() +
  ylab("AfD vote share") +
  xlab("Date") + xlim(as.Date("2015-01-01"), as.Date("2016-07-01")) +
  ggtitle("Share of respondents indicating support for AfD", "Source: Poll of Polls")

@@

The AfD's electoral prospect as reported in opinion polls is plotted in the panel below, for the same time frame. As we can see, the increase in voting intention for the radical right party trails the increase in salience by several months. In fact, the party experiences a \*low\* right around the peak in immigration coverage. Why might that be? Is it really the case that it took voters three to six months to understand that migration was a pressing issue on the political agenda, despite the omnipresence of the issue in the news media? I suggest another explanation, which points to the changing *content* of the migration discourse. While the media coverage in the early stages of the refugee movements in 2015 focused on the tragic drownings in the Mediterranean sea, reasons for refugees' escape from their home countries, and humanitarian issues, the discourse slowly shifted to focus on perceived problems with the newly arrived. Especially new years eve 2015/16 marked a turning point, when hundreds of sexual abuses by men whom the victims perceived to be of arabic descent in Cologne were reported (if you look closely, you can see a discontinuity surrounding that date in the AfD-share, which increases around one and a half percentage points at that very date).

@@

One key difference compared to other models of party competition, especially the one described in @Meguid2005 is that in the present model, party positions are considered fixed, while it is the media debate that moves. Large shifts in parties' issue positions are generally rare [@SOURCE], but the media debate is in permanent flux, promoting a constantly changing set of issues and related considerations to voters. Political actors sponsor frames and arguments, only some of which get to influence the debate. While the classic model described in @Meguid2005 views parties' electoral performance as a function of party position and issue salience, the present model views their electoral performance as a function of issue salience and the similarity of the dominant set of issue considerations to the set of considerations emphasized by the party.

@@

In order to study whether framing can change parties' electoral prospects, an ideal case would show substantial variation surrounding the framing of an issue but little variation in parties' positions. Ideally, the variation in salience would be limited and uncorrelated to the variation of the framing. I choose the case of the election campaign for the German Bundestagswahl 2017 as I consider most of these criteria are met. After the refugee movements throughout Europe in 2015, voters considered migration the most important issue [@Wessels2021, p. 12], and the media still discussed the issue, but at lower levels compared to 2017 (see figure \@ref(fig:salplot)). I expect that events during the election campaign, such as migrant crimes, conflicts regarding the intra-EU distribution of refugees, international terrorism, humanitarian crises as those among the Rohingya in Myanmar, as well as drownings in the Mediterranean sea lead to substantial variation in the set of considerations regarding migration being discussed - the key independent variable.

>>>>

---
title: "Frame Competition - Assessing the Impact of News Framing on Parties' Electoral Performance with Document Embeddings"
author: "Nicolai Berk"
date: "`r Sys.Date()`"
output:
  bookdown::pdf_document2:
    toc: false
    fig_caption: yes
    includes:
      in_header: preamble.tex
subtitle: Manuscript prepared for the OX|BER workshop, Berlin, July 14th-15th.
fontsize: 12pt
geometry: margin=1.5in
abstract: "Issue ownership theory assumes that the salience of specific issues is always more or less beneficial to a given party. I introduce the concept of frame competition and argue that a given issue can be more or less beneficial for parties, dependent on the dominant framing. I test this argument by estimating the similarity of media coverage about migration to parties' migration framing with document embeddings, using data from Germany 2013-2019. Combining this similarity measure with VAR models, I show whether higher frame similarity is associated with better performance in the polls. The findings broaden our understanding of voting behaviour and the strategies political parties can employ in the face of a hostile issue agenda."
bibliography: C:/Users/nicol/Dropbox/Studium/BibTex/PhD.bib
---

@@

knitr::opts_chunk$set(echo = F, message = F, warning = F, cache = T, fig.height=6, fig.width=10, fig.align='center', eval = T)

@@

*This manuscript has been enriched by commentary from Tarik Abou-Chadi, Daniel Bischof, Luis Cornago Bonal, Sophia Hunger, Antonio Valentim, Klaudia Wegschaider, Lisa Zehnter.*

@@

One of the most prominent explanations for parties' electoral success in scholarly and public debate is the issue agenda at a given election: the more prominent an issue was among political parties, news media, and voters, the more it would influence the latter's electoral decision making. Certain issues are therefore more favorable to given parties and parties will try to focus on 'their' issues in public communication [@Budge1983; @Petrocik1996; @Walgrave2012]. In extension, it is assumed that the main form of electoral competition in contemporary Western European democracies is *issue competition*, in the sense that parties compete to bring 'their' issue on the agenda [@GreenPedersen2019].

@@

<!-- Sophia Hunger: does associative framing really affect voting? [I am not sure if it is so relevant to go into this distinction] -->

@@

<!-- Sophia HUnger finds the above paragraph very convincing ;) -->

@@

afd_poll_plot <- 
  de_polls_df %>% 
  mutate(date = as.Date(date)) %>% 
  filter(date < as.Date("2016-07-01"),
         date >= as.Date("2015-01-01")) %>% 
  mutate(week = lubridate::floor_date(date, "week")
         , post = date >= as.Date("2016-01-01")
         ) %>% 
  group_by(week
           , post
           ) %>% 
  summarise(parties.AfD = mean(parties.AfD)) %>% 
  ggplot(aes(week, parties.AfD
             , group = post
             )) +
  geom_point() +
  geom_smooth(method = "gam") +
  # geom_vline(xintercept = as.Date("2015-09-01"), col = "red", lty = 2) +
  theme_minimal() +
  ylab("AfD vote share") +
  xlab("Date") + xlim(as.Date("2015-01-01"), as.Date("2016-07-01")) +
  ggtitle("Share of respondents indicating support for AfD", "Source: Poll of Polls")

@@

The AfD's electoral prospect as reported in opinion polls is plotted in the panel below, for the same time frame. As we can see, the increase in voting intention for the radical right party trails the increase in salience by several months. In fact, the party experiences a *low* right around the peak in immigration coverage. Why might that be? Is it really the case that it took voters three to six months to understand that migration was a pressing issue on the political agenda, despite the omnipresence of the issue in the news media? I suggest another explanation, which points to the changing *content* of the migration discourse. While the media coverage in the early stages of the refugee movements in 2015 focused on the tragic drownings in the Mediterranean sea, reasons for refugees' escape from their home countries, and humanitarian issues, the discourse slowly shifted to focus on perceived problems with the newly arrived. Especially new years eve 2015/16 marked a turning point, when hundreds of sexual abuses by men whom the victims perceived to be of arabic descent in Cologne were reported (if you look closely, you can see a discontinuity surrounding that date in the AfD-share, which increases around one and a half percentage points at that very date).

<!-- Sophia Hunger likes this -->

@@

One key difference compared to other models of party competition, especially the one described in @Meguid2005 is that in the present model, party positions are considered fixed, while it is the media debate that moves. Large shifts in parties' issue positions are generally rare [@Gessler2022], but the media debate is in permanent flux, promoting a constantly changing set of issues and related considerations to voters. Political actors sponsor frames and arguments, only some of which get to influence the debate. While the classic model described in @Meguid2005 views parties' electoral performance as a function of party position and issue salience, the present model views their electoral performance as a function of issue salience and the similarity of the dominant set of issue considerations to the set of considerations emphasized by the party.

@@

In order to study whether framing can change parties' electoral prospects, an ideal case would show substantial variation surrounding the framing of an issue but little variation in parties' positions. Ideally, the variation in salience would be limited and uncorrelated to the variation of the framing. I choose the case of the election campaign for the German Bundestagswahl 2017 as I consider most of these criteria are met. After the refugee movements throughout Europe in 2015, voters considered migration the most important issue [@Wessels2021, p. 12], and the media still discussed the issue, but at lower levels compared to 2015 (see figure \@ref(fig:salplot)). I expect that events during the election campaign, such as migrant crimes, conflicts regarding the intra-EU distribution of refugees, international terrorism, humanitarian crises as those among the Rohingya in Myanmar, as well as drownings in the Mediterranean sea lead to substantial variation in the set of considerations regarding migration being discussed - the key independent variable.
